Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome

### 2016-06-03 submission/Sze_Obesity_mBio_2016.tex
Remove Taable 1 from pdf - wasn't rendering correctly
Editing manuscript

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

@@

The recent publication of additional studies that collected BMI data for
each subject as well as other studies that were not included in the
earlier re-analysis studies offered the opportunity to revisit the
question relating the structure of the human microbiome to obesity
(14--22). One critique of the prior re-analysis studies is that the
authors did not aggregate the results across studies to increase the
effective sample size. It is possible that there were small associations
within each study that were not statistically significant because the
individual studies lacked sufficient power. Alternatively, diversity
metrics may mask the appropriate signal and it is necessary to measure
the association at the level of microbial populations. Walters et al.
(8) demonstrated that Random Forest machine learning models were capable
of predicting obesity status within a single cohort, but did not attempt
to test the models on other cohorts. The purpose of this study was to
perform a meta-analysis of the association between differences in the
microbiome and obesity status by analyzing and applying a more
systematic and synthetic approach than was used previously.

\textbf{\emph{Literature Review and Study Inclusion.}} We followed the
Preferred Reporting Items for Systematic Reviews and Meta-Analyses
(PRISMA) guidelines to identify studies to include in our meta-analysis
(23). A detailed description of our selection process and the exact
search terms are provided in the Supplement and in figure 1. Briefly, we
searched PubMed for original research studies that involved studying
obesity and the human microbiome. The initial search yielded 187
studies. We identified number\_string{[}n{]} additional studies that
were not designed to explicitly test for an association between the
microbiome and obesity. We then manually curated the 196 studies to
select those studies that included BMI and sequence data. This yielded
10 eligible studies. An additional study was removed from our analysis
because no individuals in the study had a BMI over 30. Among the final
10 studies, 3 were from identified from our PubMed search (6, 15, 20), 5
were originally identified from the 9 studies that did not explicitly
investigate obesity but included BMI data (14, 18, 19, 24, 25), and two
datasets were used (21, 22) because these publications did not
specifically look for any metabolic or obesity conditions but had
control populations and enabled us to help mitigate against publication
biases associated with the bacterial microbiome and obesity. The
number\_string{[}n{]} studies are summarized in Tables 1 and 2.

\textbf{\emph{Sequence Analysis Pipeline.}} All sequence data were
publicly available and were downloaded from the NCBI Sequence Read
Archive, the European Nucleotide Archive, or the investigators' personal
website
(\url{https://gordonlab.wustl.edu/TurnbaughSE/_10/_09/STM/_2009.html}).
In total seven studies used 454 (6, 14, 15, 19, 20, 22, 25) and three
studies used Illumina sequencing (18, 21, 24). All of these studies used
amplification-based 16S rRNA gene sequencing. Among the studies that
sequenced the 16S rRNA gene, the researchers targeted the V1-V2 (19),
V1-V3 (14, 15, 20), V3-V5 (22, 25), V4 {[}(18); (21); {]}, and V3-4 (24)
regions. For those studies where multiple regions were sequenced, we
selected the region that corresponded to the largest number of subjects
(6, 25). We processed the 16S rRNA gene sequence data using a
standardized mothur pipeline. Briefly, our pipelines attempted to follow
previously recommended approaches for 454 and Illumina sequencing data
(26, 27). All sequences were screened for chimeras using UCHIME and
assigned to operational taxonomic units (OTUs) using the average
neighbor algorithm using a 3\% distance threshold (28, 29). All sequence
processing was performed using mothur (v.1.37.0) (30).

\textbf{\emph{Data Analysis.}} We split the overall meta-analysis into
three general strategies using R (3.3.0). First, we followed the
approach employed by Finucane et al (9) and Walters et al (8) where each
study was re-analyzed separately to identify associations between BMI
and the relative abundance of Bacteroidetes and Firmicutes, the ratio of
Bacteroidetes and Firmicutes relative abundances (B:F), Shannon
diversity, observed richness, and Shannon evenness. After each variable
was transformed to fit a normal distribution a two-tailed t-test was
performed for comparison of non-obese and obese individuals (i.e.~BMI
\textgreater{} 35.0). We performed a pooled analysis on these measured
variables using linear random effect models to correct for study effect
to asses differences on the combined dataset between non-obese and obese
groups using the lme4 (v.1.1-12) R package. Next, we compared the
community structure from non-obese and obese individuals using PERMANOVA
analysis of Bray-Curtis distance matrices. This analysis was performed
using the vegan (v.2.3-5) R package. For both analyses, the datasets
were rarefied (N=1000) so that each study within a study had the same
number of sequences. Second, for each study we partitioned the subjects
into a low or high group depending on whether their alpha diversity
metrics were below or above the median value for the study. The relative
risk (RR) was then calculated as the ratio of the number of obese
individuals in the low group to the number of obese individuals in the
high group. We then performed a Fisher exact-test to investigate whether
the RR was significantly different from 1.0 within each study and across
all of the studies using the epiR (0.9-77) and metafor (1.9-8) packages.
Third, we used the AUCRF (1.1) R package to generate Random Forest
models. For each study we developed models using either OTUs or
genus-level phylotypes. The quality of each model was assessed by
measuring the area under the curve (AUC) of the Receive Operating
Characteristic (ROC) using ten-fold cross validation. Because the
genus-level phylotype models were developed using a common reference, it
was possible to use one study's model (i.e.~the training set) to
classify the samples from the other studies (i.e.~the testing sets). The
optimum threshold for the training set was set as the probability
threshold that had the highest combined sensitivity and specificity.
This threshold was then used to calculate the accuracy of the model
applied to the test studies. To generate Receiver Operator
Characteristic (ROC) curves and calculate the accuracy of the models we
used the pROC (1.8) R package. Finally, we performed power and sample
number simulations for different effect sizes for each study using the
pwr (1.1-3) R package and base R functions. We also calculated the
actual sample size needed based on the effect size of each individual
study.

@@

@@

\textbf{\emph{Beta diversity analysis.}} Following the approach used by
the Walters and Finucane re-analysis studies, for each dataset we
calculated a Bray-Curtis distance matrix to measure the difference in
the membership and structure of the individuals from each study. We then
used PERMANOVA to test for a significant differences between the
structure of non-obese and obese individuals. The Escobar, Goodrich, and
Turnbaugh datasets indicated a significant difference in community
structure (all P\textless{}0.05). Because it was not possible to
ascertain the directionality of the difference in community structure
nor perform a pooled analysis using studies that had non-overlapping 16S
rRNA gene sequence regions it is unclear whether these differences
reflect a broader, but perhaps small, shift in community structure
between non-obese and obese individuals.

@@

By selecting a range of possible effect sizes, we were able to
demonstrate that most studies are underpowered to detect modest
differences in alpha diversity metrics and phylum-level relative
abundances. Several factors interact to limit the power of microbiome
studies. There is wide interpersonal variation in the diversity and
structure of the human microbiome. In addition, the common experimental
designs limit their power. As we observed, most of the studies included
in our analysis were unbalanced for the variable that we were interested
in. This was also true of those studies that originally sought to
identify associations with obesity. Even with a balanced design, we
showed that it was necessary to obtain approximately 140 and 6,300
sequences per sample to detect a 5\% difference in Shannon diversity or
B:F, respectively. It was interesting that these sample sizes agreed
across studies regardless of their sequencing method, region with in the
16S rRNA gene, or subject population (Figure 6). This suggests that
regardless of the treatment or category, these sample sizes represent a
good starting point for subject recruitment when using stool samples.
Unfortunately, few studies have been published with this level of
subject recruitment. This is troubling since the positive predictive
rate of a significant finding in an underpowered study is small leading
to results that cannot be reproduced (31). Future microbiome studies
should articulate the basis for their experimental design.

Two previous reviews (8, 9) have stated that there was not a consistent
association between alpha diversity and obesity; however, neither of
these studies made an attempt to pool the existing data together to try
and harness the additional power that this would give and they did not
assess whether the studies were sufficiently powered to detect a
difference. Our analysis also used 16S rRNA gene sequence data from ten
studies whereas the Finucane study used 16S rRNA gene sequence data from
3 studies (7, 10, 25) and a metagenomic study (32) and the Walters study
used 16S rRNA gene sequence data from 5 studies (10, 15, 19, 25, 33);
two studies were included in both analyses (10, 25). Our analysis
included 4 of these studies (10, 15, 19, 25) and excluded 3 of the
studies because they were too small (7), only utilized metagenomic data
(32), or used short single read Illumina HiSeq data that has a high
error rate making it untractable for \emph{de novo} OTU clustering (33).
The additional seven datasets were published after the two reviews were
performed and include datasets with more samples than were found in the
original studies. Our collection of ten studies allowed us to largely
use the same sequence analysis pipeline for all datasets and relied
heavily on the availability of public data and access to metadata that
included variables beyond the needs of the original study. To execute
this analysis, we created an automated data analysis pipeline, which can
be easily updated to add additional studies as they become available.
Similarly, it would be possible to adapt this pipeline to other body
sites and treatment or variables (e.g.~subject's sex or age).

@@

@@

\textbf{Table 1. Summary Demographics of Individuals used in the
Meta-analysis.}

\textbf{Figure 1: PRISMA flow diagram of total records searched (34).}

@@

\hyperdef{}{references}{\label{references}}
\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\hyperdef{}{ref-ogdenux5fprevalenceux5f2014}{\label{ref-ogdenux5fprevalenceux5f2014}}
1. \textbf{Ogden CL}, \textbf{Carroll MD}, \textbf{Kit BK},
\textbf{Flegal KM}. 2014. Prevalence of childhood and adult obesity in
the United States, 2011-2012. JAMA \textbf{311}:806--814.
doi:\url{http://doi.org/10.1001/jama.2014.732}.

\hyperdef{}{ref-lichtashux5fbodyux5f2013}{\label{ref-lichtashux5fbodyux5f2013}}
2. \textbf{Lichtash CT}, \textbf{Cui J}, \textbf{Guo X}, \textbf{Chen
Y-DI}, \textbf{Hsueh WA}, \textbf{Rotter JI}, \textbf{Goodarzi MO}.
2013. Body adiposity index versus body mass index and other
anthropometric traits as correlates of cardiometabolic risk factors.
PloS One \textbf{8}:e65954.
doi:\url{http://doi.org/10.1371/journal.pone.0065954}.

\hyperdef{}{ref-braheux5fcanux5f2016}{\label{ref-braheux5fcanux5f2016}}
3. \textbf{Brahe LK}, \textbf{Astrup A}, \textbf{Larsen LH}. 2016. Can
We Prevent Obesity-Related Metabolic Diseases by Dietary Modulation of
the Gut Microbiota? Advances in Nutrition (Bethesda, Md)
\textbf{7}:90--101. doi:\url{http://doi.org/10.3945/an.115.010587}.

\hyperdef{}{ref-drorux5fmicrobiotaux5f2016}{\label{ref-drorux5fmicrobiotaux5f2016}}
4. \textbf{Dror T}, \textbf{Dickstein Y}, \textbf{Dubourg G},
\textbf{Paul M}. 2016. Microbiota manipulation for weight change.
Microbial Pathogenesis.
doi:\url{http://doi.org/10.1016/j.micpath.2016.01.002}.

\hyperdef{}{ref-leyux5fobesityux5f2005}{\label{ref-leyux5fobesityux5f2005}}
5. \textbf{Ley RE}, \textbf{Bäckhed F}, \textbf{Turnbaugh P},
\textbf{Lozupone CA}, \textbf{Knight RD}, \textbf{Gordon JI}. 2005.
Obesity alters gut microbial ecology. Proceedings of the National
Academy of Sciences of the United States of America
\textbf{102}:11070--11075.
doi:\url{http://doi.org/10.1073/pnas.0504978102}.

\hyperdef{}{ref-turnbaughux5fcoreux5f2009}{\label{ref-turnbaughux5fcoreux5f2009}}
6. \textbf{Turnbaugh PJ}, \textbf{Hamady M}, \textbf{Yatsunenko T},
\textbf{Cantarel BL}, \textbf{Duncan A}, \textbf{Ley RE}, \textbf{Sogin
ML}, \textbf{Jones WJ}, \textbf{Roe BA}, \textbf{Affourtit JP},
\textbf{Egholm M}, \textbf{Henrissat B}, \textbf{Heath AC},
\textbf{Knight R}, \textbf{Gordon JI}. 2009. A core gut microbiome in
obese and lean twins. Nature \textbf{457}:480--484.
doi:\url{http://doi.org/10.1038/nature07540}.

\hyperdef{}{ref-leyux5fmicrobialux5f2006}{\label{ref-leyux5fmicrobialux5f2006}}
7. \textbf{Ley RE}, \textbf{Turnbaugh PJ}, \textbf{Klein S},
\textbf{Gordon JI}. 2006. Microbial ecology: Human gut microbes
associated with obesity. Nature \textbf{444}:1022--1023.
doi:\url{http://doi.org/10.1038/4441022a}.

\hyperdef{}{ref-waltersux5fmeta-analysesux5f2014}{\label{ref-waltersux5fmeta-analysesux5f2014}}
8. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
Meta-analyses of human gut microbes associated with obesity and IBD.
FEBS letters \textbf{588}:4223--4233.
doi:\url{http://doi.org/10.1016/j.febslet.2014.09.039}.

\hyperdef{}{ref-finucaneux5ftaxonomicux5f2014}{\label{ref-finucaneux5ftaxonomicux5f2014}}
9. \textbf{Finucane MM}, \textbf{Sharpton TJ}, \textbf{Laurent TJ},
\textbf{Pollard KS}. 2014. A taxonomic signature of obesity in the
microbiome? Getting to the guts of the matter. PloS One
\textbf{9}:e84689.
doi:\url{http://doi.org/10.1371/journal.pone.0084689}.

\hyperdef{}{ref-turnbaughux5fobesity-associatedux5f2006}{\label{ref-turnbaughux5fobesity-associatedux5f2006}}
10. \textbf{Turnbaugh PJ}, \textbf{Ley RE}, \textbf{Mahowald MA},
\textbf{Magrini V}, \textbf{Mardis ER}, \textbf{Gordon JI}. 2006. An
obesity-associated gut microbiome with increased capacity for energy
harvest. Nature \textbf{444}:1027--31.
doi:\url{http://doi.org/10.1038/nature05414}.

\hyperdef{}{ref-Koren2012}{\label{ref-Koren2012}}
11. \textbf{Koren O}, \textbf{Goodrich JK}, \textbf{Cullender TC},
\textbf{Spor A}, \textbf{Laitinen K}, \textbf{Bäckhed HK},
\textbf{Gonzalez A}, \textbf{Werner JJ}, \textbf{Angenent LT},
\textbf{Knight R}, \textbf{Bäckhed F}, \textbf{Isolauri E},
\textbf{Salminen S}, \textbf{Ley RE}. 2012. Host remodeling of the gut
microbiome and metabolic changes during pregnancy. Cell
\textbf{150}:470--480.
doi:\url{http://doi.org/10.1016/j.cell.2012.07.008}.

\hyperdef{}{ref-Cox2014}{\label{ref-Cox2014}}
12. \textbf{Cox LM}, \textbf{Yamanishi S}, \textbf{Sohn J},
\textbf{Alekseyenko AV}, \textbf{Leung JM}, \textbf{Cho I}, \textbf{Kim
SG}, \textbf{Li H}, \textbf{Gao Z}, \textbf{Mahana D}, \textbf{Rodriguez
JGZ}, \textbf{Rogers AB}, \textbf{Robine N}, \textbf{Loke P},
\textbf{Blaser MJ}. 2014. Altering the intestinal microbiota during a
critical developmental window has lasting metabolic consequences. Cell
\textbf{158}:705--721.
doi:\url{http://doi.org/10.1016/j.cell.2014.05.052}.

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

@@

The recent publication of additional studies that collected BMI data for
each subject as well as other studies that were not included in the
earlier re-analysis studies offered the opportunity to revisit the
question relating the structure of the human microbiome to obesity. One
critique of the prior re-analysis studies is that the authors did not
aggregate the results across studies to increase the effective sample
size. It is possible that there were small associations within each
study that were not statistically significant because the individual
studies lacked sufficient power. Alternatively, diversity metrics may
mask the appropriate signal and it is necessary to measure the
association at the level of microbial populations. Walters et al. (8)
demonstrated that Random Forest machine learning models were capable of
predicting obesity status within a single cohort, but did not attempt to
test the models on other cohorts. The purpose of this study was to
perform a meta-analysis of the association between differences in the
microbiome and obesity status by analyzing and applying a more
systematic and synthetic approach than was used previously.

@@

\textbf{\emph{Meta-analysis}}

(10, 14--22)

@@

\textbf{\emph{Beta diversity analysis.}} Following the approach used by
the Walters and Finucane re-analysis studies, for each dataset we
calculated a Bray-Curtis distance matrix to measure the difference in
the membership and structure of the individuals from each study. We then
used AMOVA to test for a significant differences between the structure
of non-obese and obese individuals. The Escobar, Goodrich, and Turnbaugh
datasets indicated a significant difference in community structure (all
P\textless{}0.05). Because it was not possible to ascertain the
directionality of the difference in community structure nor perform a
pooled analysis using studies that had non-overlapping 16S rRNA gene
sequence regions it is unclear whether these differences reflect a
broader, but perhaps small, shift in community structure between
non-obese and obese individuals.

@@

By selecting a range of possible effect sizes, we were able to
demonstrate that most studies are underpowered to detect modest
differences in alpha diversity metrics and phylum-level relative
abundances. Several factors interact to limit the power of microbiome
studies. There is wide interpersonal variation in the diversity and
structure of the human microbiome. In addition, the common experimental
designs limit their power. As we observed, most of the studies included
in our analysis were unbalanced for the variable that we were interested
in. This was also true of those studies that originally sought to
identify associations with obesity. Even with a balanced design, we
showed that it was necessary to obtain approximately 140 and 6,300
sequences per sample to detect a 5\% difference in Shannon diversity or
B:F, respectively. It was interesting that these sample sizes agreed
across studies regardless of their sequencing method, region with in the
16S rRNA gene, or subject population (Figure 6). This suggests that
regardless of the treatment or category, these sample sizes represent a
good starting point for subject recruitment when using stool samples.
Unfortunately, few studies have been published with this level of
subject recruitment. This is troubling since the positive predictive
rate of a significant finding in an underpowered study is small leading
to results that cannot be reproduced (23). Future microbiome studies
should articulate the basis for their experimental design.

Two previous reviews (8, 9) have stated that there was not a consistent
association between alpha diversity and obesity; however, neither of
these studies made an attempt to pool the existing data together to try
and harness the additional power that this would give and they did not
assess whether the studies were sufficiently powered to detect a
difference. Our analysis also used 16S rRNA gene sequence data from ten
studies whereas the Finucane study used 16S rRNA gene sequence data from
3 studies (7, 10, 22) and a metagenomic study (24) and the Walters study
used 16S rRNA gene sequence data from 5 studies (10, 15, 17, 22, 25);
two studies were included in both analyses (10, 22). Our analysis
included 4 of these studies (10, 15, 17, 22) and excluded 3 of the
studies because they were too small (7), only utilized metagenomic data
(24), or used short single read Illumina HiSeq data that has a high
error rate making it untractable for \emph{de novo} OTU clustering (25).
The additional seven datasets were published after the two reviews were
performed and include datasets with more samples than were found in the
original studies. Our collection of ten studies allowed us to largely
use the same sequence analysis pipeline for all datasets and relied
heavily on the availability of public data and access to metadata that
included variables beyond the needs of the original study. To execute
this analysis, we created an automated data analysis pipeline, which can
be easily updated to add additional studies as they become available.
Similarly, it would be possible to adapt this pipeline to other body
sites and treatment or variables (e.g.~subject's sex or age).

@@

\textbf{\emph{Literature Review and Study Inclusion.}} We followed the
Preferred Reporting Items for Systematic Reviews and Meta-Analyses
(PRISMA) guidelines to identify studies to include in our meta-analysis
(26). A detailed description of our selection process and the exact
search terms are provided in the Supplement and in figure 1. Briefly, we
searched PubMed for original research studies that involved studying
obesity and the human microbiome. The initial search yielded 187
studies. We identified number\_string{[}n{]} additional studies that
were not designed to explicitly test for an association between the
microbiome and obesity. We then manually curated the 196 studies to
select those studies that included BMI and sequence data. This yielded
10 eligible studies. An additional study was removed from our analysis
because no individuals in the study had a BMI over 30. Among the final
10 studies, 3 were from identified from our PubMed search (6, 15, 18), 5
were originally identified from the 9 studies that did not explicitly
investigate obesity but included BMI data (14, 16, 17, 21, 22), and two
datasets were used (19, 20) because these publications did not
specifically look for any metabolic or obesity conditions but had
control populations and enabled us to help mitigate against publication
biases associated with the bacterial microbiome and obesity. The
number\_string{[}n{]} studies are summarized in Tables 1 and 2.

\textbf{\emph{Sequence Analysis Pipeline.}} All sequence data were
publicly available and were downloaded from the NCBI Sequence Read
Archive, the European Nucleotide Archive, or the investigators' personal
website
(\url{https://gordonlab.wustl.edu/TurnbaughSE/_10/_09/STM/_2009.html}).
In total seven studies used 454 (6, 14, 15, 17, 18, 20, 22) and three
studies used Illumina sequencing (16, 19, 21). All of these studies used
amplification-based 16S rRNA gene sequencing. Among the studies that
sequenced the 16S rRNA gene, the researchers targeted the V1-V2 (17),
V1-V3 (14, 15, 18), V3-V5 (20, 22), V4 {[}(16); (19); {]}, and V3-4 (21)
regions. For those studies where multiple regions were sequenced, we
selected the region that corresponded to the largest number of subjects
(6, 22). We processed the 16S rRNA gene sequence data using a
standardized mothur pipeline. Briefly, our pipelines attempted to follow
previously recommended approaches for 454 and Illumina sequencing data
(27, 28). All sequences were screened for chimeras using UCHIME and
assigned to operational taxonomic units (OTUs) using the average
neighbor algorithm using a 3\% distance threshold (29, 30). All sequence
processing was performed using mothur (v.1.37.0) (31).

\textbf{\emph{Data Analysis.}} We split the overall meta-analysis into
three general strategies using R (3.3.0). First, we followed the
approach employed by Finucane et al (9) and Walters et al (8) where each
study was re-analyzed separately to identify associations between BMI
and the relative abundance of Bacteroidetes and Firmicutes, the ratio of
Bacteroidetes and Firmicutes relative abundances (B:F), Shannon
diversity, observed richness, and Shannon evenness. After each variable
was transformed to fit a normal distribution a two-tailed t-test was
performed for comparison of non-obese and obese individuals (i.e.~BMI
\textgreater{} 35.0). We performed a pooled analysis on these measured
variables using linear random effect models to correct for study effect
to asses differences on the combined dataset between non-obese and obese
groups using the lme4 (v.1.1-12) R package. Next, we compared the
community structure from non-obese and obese individuals using analysis
of molecular variance (AMOVA) with Bray-Curtis distance matrices (32).
This analysis was performed using the vegan (v.2.3-5) R package. For
both analyses, the datasets were rarefied (N=1000) so that each study
within a study had the same number of sequences. Second, for each study
we partitioned the subjects into a low or high group depending on
whether their alpha diversity metrics were below or above the median
value for the study. The relative risk (RR) was then calculated as the
ratio of the number of obese individuals in the low group to the number
of obese individuals in the high group. We then performed a Fisher
exact-test to investigate whether the RR was significantly different
from 1.0 within each study and across all of the studies using the epiR
(0.9-77) and metafor (1.9-8) packages. Third, we used the AUCRF (1.1) R
package to generate Random Forest models. For each study we developed
models using either OTUs or genus-level phylotypes. The quality of each
model was assessed by measuring the area under the curve (AUC) of the
Receive Operating Characteristic (ROC) using ten-fold cross validation.
Because the genus-level phylotype models were developed using a common
reference, it was possible to use one study's model (i.e.~the training
set) to classify the samples from the other studies (i.e.~the testing
sets). The optimum threshold for the training set was set as the
probability threshold that had the highest combined sensitivity and
specificity. This threshold was then used to calculate the accuracy of
the model applied to the test studies. To generate Receiver Operator
Characteristic (ROC) curves and calculate the accuracy of the models we
used the pROC (1.8) R package. Finally, we performed power and sample
number simulations for different effect sizes for each study using the
pwr (1.1-3) R package and base R functions. We also calculated the
actual sample size needed based on the effect size of each individual
study.

@@

\textbf{Figure 1: PRISMA flow diagram of total records searched (33).}

@@

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\hypertarget{ref-ogdenux5fprevalenceux5f2014}{}
1. \textbf{Ogden CL}, \textbf{Carroll MD}, \textbf{Kit BK},
\textbf{Flegal KM}. 2014. Prevalence of childhood and adult obesity in
the United States, 2011-2012. JAMA \textbf{311}:806--814.
doi:\href{https://doi.org/10.1001/jama.2014.732}{10.1001/jama.2014.732}.

\hypertarget{ref-lichtashux5fbodyux5f2013}{}
2. \textbf{Lichtash CT}, \textbf{Cui J}, \textbf{Guo X}, \textbf{Chen
Y-DI}, \textbf{Hsueh WA}, \textbf{Rotter JI}, \textbf{Goodarzi MO}.
2013. Body adiposity index versus body mass index and other
anthropometric traits as correlates of cardiometabolic risk factors.
PloS One \textbf{8}:e65954.
doi:\href{https://doi.org/10.1371/journal.pone.0065954}{10.1371/journal.pone.0065954}.

\hypertarget{ref-braheux5fcanux5f2016}{}
3. \textbf{Brahe LK}, \textbf{Astrup A}, \textbf{Larsen LH}. 2016. Can
We Prevent Obesity-Related Metabolic Diseases by Dietary Modulation of
the Gut Microbiota? Advances in Nutrition (Bethesda, Md)
\textbf{7}:90--101.
doi:\href{https://doi.org/10.3945/an.115.010587}{10.3945/an.115.010587}.

\hypertarget{ref-drorux5fmicrobiotaux5f2016}{}
4. \textbf{Dror T}, \textbf{Dickstein Y}, \textbf{Dubourg G},
\textbf{Paul M}. 2016. Microbiota manipulation for weight change.
Microbial Pathogenesis.
doi:\href{https://doi.org/10.1016/j.micpath.2016.01.002}{10.1016/j.micpath.2016.01.002}.

\hypertarget{ref-leyux5fobesityux5f2005}{}
5. \textbf{Ley RE}, \textbf{Bäckhed F}, \textbf{Turnbaugh P},
\textbf{Lozupone CA}, \textbf{Knight RD}, \textbf{Gordon JI}. 2005.
Obesity alters gut microbial ecology. Proceedings of the National
Academy of Sciences of the United States of America
\textbf{102}:11070--11075.
doi:\href{https://doi.org/10.1073/pnas.0504978102}{10.1073/pnas.0504978102}.

\hypertarget{ref-turnbaughux5fcoreux5f2009}{}
6. \textbf{Turnbaugh PJ}, \textbf{Hamady M}, \textbf{Yatsunenko T},
\textbf{Cantarel BL}, \textbf{Duncan A}, \textbf{Ley RE}, \textbf{Sogin
ML}, \textbf{Jones WJ}, \textbf{Roe BA}, \textbf{Affourtit JP},
\textbf{Egholm M}, \textbf{Henrissat B}, \textbf{Heath AC},
\textbf{Knight R}, \textbf{Gordon JI}. 2009. A core gut microbiome in
obese and lean twins. Nature \textbf{457}:480--484.
doi:\href{https://doi.org/10.1038/nature07540}{10.1038/nature07540}.

\hypertarget{ref-leyux5fmicrobialux5f2006}{}
7. \textbf{Ley RE}, \textbf{Turnbaugh PJ}, \textbf{Klein S},
\textbf{Gordon JI}. 2006. Microbial ecology: Human gut microbes
associated with obesity. Nature \textbf{444}:1022--1023.
doi:\href{https://doi.org/10.1038/4441022a}{10.1038/4441022a}.

\hypertarget{ref-waltersux5fmeta-analysesux5f2014}{}
8. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
Meta-analyses of human gut microbes associated with obesity and IBD.
FEBS letters \textbf{588}:4223--4233.
doi:\href{https://doi.org/10.1016/j.febslet.2014.09.039}{10.1016/j.febslet.2014.09.039}.

\hypertarget{ref-finucaneux5ftaxonomicux5f2014}{}
9. \textbf{Finucane MM}, \textbf{Sharpton TJ}, \textbf{Laurent TJ},
\textbf{Pollard KS}. 2014. A taxonomic signature of obesity in the
microbiome? Getting to the guts of the matter. PloS One
\textbf{9}:e84689.
doi:\href{https://doi.org/10.1371/journal.pone.0084689}{10.1371/journal.pone.0084689}.

\hypertarget{ref-turnbaughux5fobesity-associatedux5f2006}{}
10. \textbf{Turnbaugh PJ}, \textbf{Ley RE}, \textbf{Mahowald MA},
\textbf{Magrini V}, \textbf{Mardis ER}, \textbf{Gordon JI}. 2006. An
obesity-associated gut microbiome with increased capacity for energy
harvest. Nature \textbf{444}:1027--31.
doi:\href{https://doi.org/10.1038/nature05414}{10.1038/nature05414}.

\hypertarget{ref-Koren2012}{}
11. \textbf{Koren O}, \textbf{Goodrich JK}, \textbf{Cullender TC},
\textbf{Spor A}, \textbf{Laitinen K}, \textbf{Bäckhed HK},
\textbf{Gonzalez A}, \textbf{Werner JJ}, \textbf{Angenent LT},
\textbf{Knight R}, \textbf{Bäckhed F}, \textbf{Isolauri E},
\textbf{Salminen S}, \textbf{Ley RE}. 2012. Host remodeling of the gut
microbiome and metabolic changes during pregnancy. Cell
\textbf{150}:470--480.
doi:\href{https://doi.org/10.1016/j.cell.2012.07.008}{10.1016/j.cell.2012.07.008}.

\hypertarget{ref-Cox2014}{}
12. \textbf{Cox LM}, \textbf{Yamanishi S}, \textbf{Sohn J},
\textbf{Alekseyenko AV}, \textbf{Leung JM}, \textbf{Cho I}, \textbf{Kim
SG}, \textbf{Li H}, \textbf{Gao Z}, \textbf{Mahana D}, \textbf{Rodriguez
JGZ}, \textbf{Rogers AB}, \textbf{Robine N}, \textbf{Loke P},
\textbf{Blaser MJ}. 2014. Altering the intestinal microbiota during a
critical developmental window has lasting metabolic consequences. Cell
\textbf{158}:705--721.
doi:\href{https://doi.org/10.1016/j.cell.2014.05.052}{10.1016/j.cell.2014.05.052}.
